---
title: "Data Import"
author: "Anil Bashetty"
date: "April 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data import

## Loading libraries

```{r results='hide',message=FALSE}
library("tidyverse")
```

## Reading CSV files 

### Reading csv file from the place where file is located by using read_csv() function

```{r eval=FALSE}
week5 <- read_csv("D:/Harrisburg University/Spring-2019/Exploratory Data Analysis/week-5/week5.csv")
# Prints out a schema that gives the name and type of each column
```

# Create an inline csv file
# read_csv function takes the first line as column names, if not specified
```{r}
read_csv("A,B,C 
          a,b,c
          1,2,3")
```

# If we use skip = n, then it will skip the first n lines of metadata;
```{r}
read_csv("A,B,C 
          a,b,c
          1,2,3", skip = 1)
```

# If data does not have column names, then set col_names = FALSE in read_csv() to auto-name columns
```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE)
```

# If you want to set our own column names
```{r}
read_csv("1,2,3\n4,5,6", col_names = c("col1", "col2", "col3"))
```

# Option na - It is used to specify missing values in the csv file
```{r}
read_csv("1,2,3\n4,5,.", na = ".")
```

## CSV Exercise
### What function would you use to read a file where fields were separated with "|"?
```{r eval=FALSE}
read_delim(file, delim = "|")
```

### Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?
```{r}
union(names(formals(read_csv)), names(formals(read_tsv)))
```

### What are the most important arguments to read_fwf()? This function is used for col_positions, it specifies where data columns begin and end.

### Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By convention, read_csv() assumes that the quoting character will be ", and if you want to change it you'll need to use read_delim() instead. What arguments do you need to specify to read the following text into a data frame?
### x,y\n1,'a,b'" 
```{r}
x <- "x,y\n1,'a,b'"
read_delim(x, ",", quote = "'")
read_csv(x, quote = "'")

```

### Identify what is wrong with each of the following inline CSV files. What happens when you run the code?
#### Last column is dropped with ony 2 columns in header

```{r}
read_csv("a,b\n1,2,3\n4,5,6")

```

#### Row 1 - column c is missed
#### Row 2 - Last value is dropped
```{r}
read_csv("a,b,c\n1,2\n1,2,3,4")

```

#### Field separator used as ";" instead of ","
```{r}
read_csv("a;b\n1;3")

```

## Parsing a vector

### Below functions take a character vector and return a more specific vector like a logical, integer, or date

```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))

str(parse_integer(c("1", "2", "3")))

str(parse_date(c("2010-01-01", "1979-10-14")))

```

### Parsing Numbers

```{r}
parse_double("1.23")

parse_double("1,23", locale = locale(decimal_mark = ","))

parse_number("$100")

parse_number("20%")

parse_number("It cost $123.45")

```

### Parsing a string
#### Getting the underlying representation of a string using charToRaw()
#### Specifing the character encoding type in parse_character() function
```{r}
charToRaw("Hadley")
x1 <- "El Ni\xf1o was particularly bad this year"
parse_character(x1, locale = locale(encoding = "Latin1"))

```

### Parsing Factors
#### Use parse_factor() a vector of known levels to generate a warning whenever an unexpected value is present
```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)

```

### Parsing Dates / Date-Times / Times

```{r}
#### If time is omitted, it will be set to midnight
parse_datetime("2010-10-01T2010")

parse_datetime("20101010")

#### parse_date() expects a four digit year, a - or /, the month, a - or /, then the day
parse_date("2010-10-01")

#### parse_time() expects the hour, :, minutes, optionally : and seconds, and an optional am/pm specifier
parse_time("01:10 am")
parse_time("20:10:01")

```
### Writing to a file

```{r eval=FALSE}
write_csv(challenge, "challenge.csv")
#### Note always, it uses UTF-8 encoding

```

### Uniform Wrappers

```{r eval=FALSE}
#### write_rds() and read_rds() are uniform wrappers around the base functions readRDS() and saveRDS()

write_rds(challenge, "challenge.rds")
read_rds("challenge.rds")

```

### Feather Package

```{r eval=FALSE}
#### feather package implements a fast binary file format that can be shared across programming languages

write_feather(challenge, "challenge.feather")
read_feather("challenge.feather")

```
